---
title: "rdviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rdviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


`rdviz` is an `R` package that creates visualizations for regression discontinuity analyses. It builds on the automated bandwidth selection and local polynomial estimation techniques from the `rdrobust` package (Calonico, Cattaneo, Farrell and Titiunik 2015). 

## Installation

The `rdviz` package is currently available on GitHub. You can install it using the `devtools` package.

```{r installation, message=FALSE, warning=FALSE}
devtools::install_github('joeornstein/rdviz')
library(rdviz)
```

## Example

The `X` and `Y` variables are randomly generated data included with the package. Here are two examples visualizations using these data, varying the polynomial order.

```{r setup, message=FALSE, warning=FALSE}
rdviz(x = X, y = Y, c = 0, p = 1)

rdviz(x = X, y = Y, c = 0, p = 2)
```
