---
title: "COVID-19 Animation"
author: "Joseph T. Ornstein"
date: '2020-06-16'
slug: covid-animation
categories: 
 - R
 - epidemiology
tags: []
images: []
---

The `gganimate` package facets a `ggplot` by some variable and stitches the plots together to create an animated video. Here's some code to generate such an animation, using the time series of COVID-19 deaths in US counties from the *New York Times* [data repository](https://github.com/nytimes/covid-19-data).

First we load and clean up the data, merging it with county-level population counts and demographics.

```{r Load Data, echo=TRUE, message=FALSE, warning=FALSE}

library(tidyverse) 
library(sf)
library(gganimate)

# Read from NYT github
data <- read_csv('https://github.com/nytimes/covid-19-data/raw/master/us-counties.csv')

# If unavailable, use the file pulled previously
if(!exists('data')){
  data <- read_csv('data/nyt-covid/us-counties.csv')
}

# Generate new cases / deaths variables
data <- data %>% 
  group_by(county, state) %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths)) %>% 
  mutate(new_cases = if_else(is.na(new_cases), 0, new_cases),
         new_deaths = if_else(is.na(new_deaths), 0, new_deaths))

# Merge with county population / demographics
library(readxl)
countypops <- read_excel('data/census/countypop.xlsx') %>% 
  mutate(fips = substr(id, 10, 14))

race <- read_excel('data/census/countyrace.xlsx') %>% 
  mutate(fips = substr(id, 10,14))

county_data <- countypops %>% 
  left_join(race %>% select(fips, total_black), by = 'fips') %>% 
  mutate(pct_black = total_black / population * 100)

data <- data %>% 
  left_join(county_data %>% 
              select(-county, -id), 
            by = 'fips') 

# Fix New York City population
nyc <- countypops %>% 
  filter(county %in% c('Kings County, New York',
                       'Queens County, New York',
                       'New York County, New York',
                       'Bronx County, New York',
                       'Richmond County, New York'))
data[data$county == 'New York City',]$population <- sum(nyc$population)

# Give New York City the Manhattan FIPS code
data[data$county == 'New York City',]$fips <- '36061'
```

Next, get shapefiles for the continental US from the `tigris` package.

```{r Load Shapefiles, warning=FALSE, message=FALSE}

continental_states <- unique(tigris::fips_codes$state)[c(1, 3:11, 13:51)]

state_boundaries <- tigris::states(class = 'sf', cb = TRUE, progress_bar = FALSE) %>%
  filter(STUSPS %in% continental_states)

county_shp <- tigris::counties(state = continental_states,
                               cb = TRUE,
                               class = 'sf',
                               progress_bar = FALSE) %>% 
  mutate(fips = paste0(STATEFP, COUNTYFP)) %>% 
  select(fips)

# add centroids (https://www.r-spatial.org/r/2018/10/25/ggplot2-sf-2.html)
county_shp <- cbind(county_shp, st_coordinates(st_centroid(county_shp))) 

county_centroids <- county_shp %>% 
  as_tibble %>% 
  select(fips, X, Y)

# merge with county data
county_shp <- county_shp %>% 
  left_join(county_data, by = 'fips')

theme_set(theme_void() +
          theme(axis.text = element_blank()))

p <- ggplot(data = county_shp) +
  geom_sf(mapping = aes(fill = pct_black)) +
  geom_point(mapping = aes(x=X,y=Y), 
             size = 0.1) +
  xlab('') + ylab('') +
  scale_fill_gradient(low = 'white', high = 'black')

p
```

Then create a `ggplot2` object with all of the time series data plotted on a map and animate it with the `gganimate::transition_states()` function.

```{r Create ggplot object and animate, warning=FALSE, message=FALSE}

p <- ggplot(data = data %>%
              left_join(county_centroids, by = 'fips') %>%
              filter(new_deaths > 0)) +
  geom_sf(data = county_shp, aes(fill = pct_black), color = NA) +
  geom_sf(data = state_boundaries, fill = NA) +
  scale_fill_gradient(low = 'white', high = 'black') +
  geom_point(mapping = aes(x=X,y=Y, size = new_deaths),
             color = 'red') +
  xlab('') + ylab('') +
  theme(legend.position = "none")

animation <- p + 
  transition_states(date, transition_length = 1,
                    state_length = 3) +
  ggtitle('{closest_state}') +
  enter_fade() + exit_fade()

# render the animation (commented out because Hugo doesn't like multiple gifs per page)
# animate(animation, nframes = 250) 

```

Animating the raw death statistics comes out a bit jerky. There's a lot of idiosyncratic day-to-day noise in the death reporting, which we can smooth out like so, complete with clunky `for` loop:


```{r smooth video, warning=FALSE, message=FALSE}

# Fit LOESS (span parameter = 0.3) for each county and append predictions
data <- data %>% 
  ungroup %>% 
  filter(!is.na(fips)) %>% 
  mutate(ID = 1:n(),
         date_numeric = as.numeric(date),
         cases_smoothed = NA,
         death_smoothed = NA)

for(i in unique(data$fips)){
  
  data_subset <- data %>% filter(fips == i)
  
  if(nrow(data_subset) < 7) next # don't smooth time series that are too short
  
  death_loess <- loess(new_deaths ~ date_numeric, 
                       data = data_subset,
                       span = 0.3)
  
  cases_loess <- loess(new_cases ~ date_numeric,
                       data = data_subset,
                       span = 0.3)
  
  data$death_smoothed[data$ID %in% data_subset$ID] <- 
    death_loess %>% 
    predict(data_subset)
  
  data$cases_smoothed[data$ID %in% data_subset$ID] <- 
    cases_loess %>% 
    predict(data_subset)
  
  
}

p <- ggplot(data = data %>%
              left_join(county_centroids, by = 'fips') %>%
              filter(death_smoothed >= 0.5)) +
  geom_sf(data = county_shp, aes(fill = pct_black), color = NA) +
  geom_sf(data = state_boundaries, fill = NA) +
  scale_fill_gradient(low = 'white', high = 'black') +
  geom_point(mapping = aes(x=X,y=Y, size = death_smoothed),
             color = 'red') +
  xlab('') + ylab('') +
  theme(legend.position = "none")

animation <- p + 
  transition_states(date, transition_length = 1,
                    state_length = 3) +
  ggtitle('{closest_state}') +
  enter_fade() + exit_fade()

# render the animation
animate(animation, nframes = 250) 


```

Much better. I have a few ideas for how to clean it up and make it more beautiful, but they will have to wait for [another post](https://joeornstein.github.io/covid-animation-2).


## Futher Reading

[`gganimate` with a Spooky Twist](https://goodekat.github.io/presentations/2019-isugg-gganimate-spooky/slides.html#1)
