% Thanks to Chris Paciorek for the template.
% https://www.stat.berkeley.edu/~paciorek/computingTips/Latex_template_creating_CV_.html

\documentclass[margin,line]{res}
\usepackage{hyperref}

% The following metadata will show up in the PDF properties
\hypersetup{
	colorlinks = true,
	urlcolor = black,
	%pdfauthor = {\name},
	pdfkeywords = {political science, cv},
	%pdftitle = {\name: Curriculum Vitae},
	pdfsubject = {Curriculum Vitae},
	pdfpagemode = UseNone
}

\oddsidemargin -.5in
\evensidemargin -.5in
\textwidth=6in
\itemsep=0in
\parsep=0in
% if using pdflatex:
\setlength{\pdfpagewidth}{\paperwidth}
\setlength{\pdfpageheight}{\paperheight} 

\newenvironment{list1}{
  \begin{list}{}{%{\ding{113}}{%
      \setlength{\itemsep}{0.1in}
      \setlength{\parsep}{0in} \setlength{\parskip}{0in}
      \setlength{\topsep}{0.1in} \setlength{\partopsep}{0in} 
      \setlength{\leftmargin}{0.17in}}}{\end{list}}
\newenvironment{list2}{
  \begin{list}{$\bullet$}{%
      \setlength{\itemsep}{0in}
      \setlength{\parsep}{0in} \setlength{\parskip}{0in}
      \setlength{\topsep}{0in} \setlength{\partopsep}{0in} 
      \setlength{\leftmargin}{0.2in}}}{\end{list}}
\newenvironment{publist}{
	\begin{list}{}{%{\ding{113}}{%
			\setlength{\itemsep}{0.15in}
			\setlength{\parsep}{0in} \setlength{\parskip}{0in}
			\setlength{\topsep}{0in} \setlength{\partopsep}{0in} 
			\setlength{\leftmargin}{0.15in}
			\setlength{\itemindent}{-0.15in}}}
		{\end{list}}

\begin{document}

\name{Joseph T. Ornstein \vspace*{.1in}}

\begin{resume}
	
\section{\sc Contact Information}
\vspace{.05in}
\begin{tabular}{@{}p{2in}p{4in}}   
Baldwin Hall 304C & Email: \href{mailto:jornstein@uga.edu}{\tt jornstein@uga.edu}\\       
University of Georgia & Website: \href{https://joeornstein.github.io/}{\tt https://joeornstein.github.io/}\\
\end{tabular}


\section{\sc Academic Positions}

{\bf University of Georgia}, Athens, Georgia
\begin{list1}
	\item[] Assistant Professor, Department of Political Science \hfill {\bf 2020 - present}
\end{list1}

{\bf Washington University in St. Louis}, St. Louis, Missouri
\begin{list1}
	\item[] Postdoctoral Research Associate, Brown School \hfill {\bf 2018 - 2020}
\end{list1}

{\bf The Brookings Institution}, Washington, DC
\begin{list1}
	\item[] Senior Research Assistant, Economic Studies \hfill {\bf 2010 - 2013}
\end{list1}



\section{\sc Education}
{\bf University of Michigan}, Ann Arbor, Michigan USA
\begin{list1}
\item[] Ph.D., Political Science, 2018
%\begin{list2}
%\vspace*{.05in}
%\item Dissertation Committee:  Robert Franzese (chair), Jenna Bednar, Elisabeth Gerber, Scott Page
%\end{list2}
\end{list1}

{\bf Dartmouth College}, Hanover, New Hampshire
\begin{list1}
\item[] B.A., Mathematics \& Social Sciences, 2009
\end{list1}




\section{\sc Peer-Reviewed Publications}

\begin{publist}
	
	\item Who Represents the Renters? How Homeownership Shapes American Politics (with Katherine Einstein \& Maxwell Palmer). 2022. \textit{Housing Policy Debate} (forthcoming). 
	
	\item \href{https://www.sciencedirect.com/science/article/pii/S1353829222000764}{Draining the Tobacco Swamps: Shaping the Built Environment to Reduce Tobacco Retailer Proximity to Residents in 30 Big US Cities (with Todd B. Combs, et al.). 2022. \textit{Health \& Place} 75}.
	
	\item \href{https://onlinelibrary.wiley.com/doi/abs/10.1111/ecpo.12204}{The Interest Premium for Left Government: Regression-Discontinuity Estimates (with Jude C. Hays and Robert J. Franzese). 2022. \textit{Economics \& Politics}}.
			
	\item \href{https://link.springer.com/article/10.1186/s13012-020-01028-5}{Rugged Landscapes: Complexity and Implementation Science (with Ross C. Brownson, Ross A. Hammond, Margaret Padek, and Stephanie Mazzucca). 2020. \textit{Implementation Science} 15(1):1-9.}
	
	\item \href{https://doi.org/10.1017/pan.2019.43}{Stacked Regression and Poststratification. 2020. \textit{Political Analysis} 28(2):293-301.}
	
	\item \href{http://dx.doi.org/10.1136/tobaccocontrol-2019-054986}{Modeling the Impact of Menthol Sales Restrictions \& Retailer Density Reduction Policies: Insights from Tobacco Town Minnesota. (with Todd B. Combs, Virginia McKay et al.). 2020. \textit{Tobacco Control} (forthcoming).}
	
	\item \href{https://joeornstein.github.io/publications/Ornstein-Hammond2017.pdf}{The Burglary Boost: A Note on Detecting Contagion Using the Knox Test (with Ross A. Hammond). 2017. \textit{Journal of Quantitative Criminology} 33(1): 65-75.}
	
	\item \href{https://joeornstein.github.io/publications/Ornstein-Norman-2014/Ornstein-Norman-2014.pdf}{Frequency of Monotonicity Failure under Instant Runoff Voting: Estimates Based on a Spatial Model of Elections (with Robert Z. Norman). 2014. \textit{Public Choice} 161(1-2): 1-9.}
	
	\item \href{https://joeornstein.github.io/publications/Hammond_et_al-2014-Annals_of_the_New_York_Academy_of_Sciences.pdf}{A Model of Social Influence on Body Mass Index (with Ross A. Hammond). 2014. \textit{Annals of the New York Academy of Sciences} 1331(1): 34-42.}
	
	\item \href{https://joeornstein.github.io/publications/Hammond-2012.pdf}{A Model of Food Reward Learning with Dynamic Reward Exposure (with Ross A. Hammond et al.). 2012. \textit{Frontiers in Computational Neuroscience} 6:82.}	
\end{publist}



%\newpage

\section{\sc Other Publications}

\begin{publist}
	\item ``Getting the Most Out of Surveys: Multilevel Regression And Poststratification''. Chapter in \textit{Detecting Causation} (ed. Fedra Negri and Alessia Damonte), Springer Nature, forthcoming.
	\item Agent-Based Models in the Social Sciences (Chapter 4) and Agent-Based Models in Public Health (Chapter 5), In \textit{New Horizons in Modeling and Simulation for Social Epidemiology and Public Health} (ed. Daniel Kim). Wiley Press, 2021. 
	\item Moving From Metrics to Mechanisms to Evaluate Tobacco Retailer Policies: Importance of Retail Policy in Tobacco Control (with Douglas A. Luke, Todd B. Combs, Lisa Henriksen, and Maggie Mahoney). 2020. \textit{American Journal of Public Health} 110(4):431-433 (invited editorial).
\end{publist}



\section{\sc Working Papers}

\begin{publist}
	\item How to Train Your Stochastic Parrot: Deep Language Models for Political Texts (with Elise Blasingame and Jake Truscott).
	\item The Trump Effect on Immigration Court Cases (with Elise Blasingame, Christina Boyd, and Roberto Carlos).
	\item Gaussian Process Regression Discontinuity (with JBrandon Duck-Mayr).
	\item \href{https://joeornstein.github.io/publications/ornstein-zone-defense.pdf}{Zone Defense: Why Liberal Cities Build Fewer Houses.}
	\item \href{https://joeornstein.github.io/publications/Ornstein-ElectionTiming.pdf}{Election Timing and The Politics of Urban Growth.}
\end{publist}

\section{\sc Conference \\and Invited Presentations}

\textbf{2022:} ReSPoS Summer School (University of Milan), Society for Political Methodology (PolMeth), American Political Science Association (APSA)

\textbf{2021:} Society for Political Methodology (PolMeth)

\textbf{2020:} International Methods Colloquium, MRP Reading Group

\textbf{2019:} WashU Network for Dissemination and Implementation Research (WUNDIR), Society for Political Methodology (PolMeth), WashU Data Lab, APSA Local Political Economy Pre-Conference, American Political Science Association

\textbf{2017:} Midwest Political Science Association, APSA Local Political Economy Pre-Conference, American Political Science Association

\textbf{2016:} Midwest Political Science Association

\textbf{2014:} Public Choice Society

\textbf{2010:} Public Choice Society


\section{\sc Teaching}


\textbf{University of Georgia}
\begin{list1}
	\item[] \href{https://joeornstein.github.io/text-as-data//}{POLS 8500: Text As Data (graduate)}
	\item[] POLS 8500: Bayesian Statistics (graduate)
	\item[] \href{https://joeornstein.github.io/pols-7012//}{POLS 7012: Introduction to Political Methodology (graduate)}
	\item[] \href{https://joeornstein.github.io/pols-4641/}{POLS 4641: The Science of Cities}
	\item[] \href{https://joeornstein.github.io/pols-3230/}{POLS 3230: Political Analysis in \texttt{R}}
	\item[] POLS 3220: Introduction to Models in Political Science
\end{list1}

\textbf{Washington University in St. Louis}
\begin{list1}
	\item[] Benefit-Cost Analysis
	\item[] Agent-Based Modeling (w/ Ross Hammond)
\end{list1}

\textbf{University of Michigan}
\begin{list1}
	\item[] The Political Economy of Cities
	\item[] Political Science Graduate Student Math Camp
	\item International Economic Relations (w/ Andrew Kerner)
	\item Introduction to Models (w/ Scott Page)
	\item Advanced Empirical Modeling for Theory Evaluation (w/ Robert Franzese)
	\item Spatial Econometrics Workshop (w/ Robert Franzese)
	\item Introduction to Comparative Politics (w/ Robert Franzese)
	\item Introduction to World Politics (w/ James Morrow)
	\item Complex Systems Modeling for Public Health Research (w/ Ross Hammond)
\end{list1}



\section{\sc Honors and Awards} 

ProQuest Distinguished Dissertation Award, Honorable Mention (2019)

Rackham One-Term Dissertation Fellowship (2018)

University of Michigan Library Data Grant (2017)

Rackham Conference Travel Grant (2013, 2015, 2016, 2017)

Dartmouth College Neukom Scholarship (2009)



\section{\sc Reviewer For}
\textit{American Journal of Political Science, Journal of Theoretical Politics, Public Choice, Political Science Research and Methods, Political Behavior, Pediatric Obesity}

\section{\sc Professional Affiliations}
American Political Science Association, Society for Political Methodology





\section{\sc Miscellaneous}
\begin{list2}
	\item \textbf{Software Proficiencies}: Stata, {\tt R}, Python, Java, NetLogo, \LaTeX
	\item \textbf{Erd{\H o}s Number}: 3
\end{list2}


\section{\sc Last Updated}
\today

%\bigskip
%
%% Footer
%\begin{center}
%	\begin{footnotesize}
%		Last updated: \today \\
%		\href{https://joeornstein.github.io/}{\tt https://joeornstein.github.io/}
%	\end{footnotesize}
%\end{center}

\end{resume}
\end{document}




