% LaTeX Curriculum Vitae Template
%
% Copyright (C) 2004-2009 Jason Blevins <jrblevin@sdf.lonestar.org>
% http://jblevins.org/projects/cv-template/
%
% You may use use this document as a template to create your own CV
% and you may redistribute the source code freely. No attribution is
% required in any resulting documents. I do ask that you please leave
% this notice and the above URL in the source code if you choose to
% redistribute this file.

\documentclass[letterpaper]{article}

\usepackage{hyperref}
\usepackage{geometry}

% Comment the following lines to use the default Computer Modern font
% instead of the Palatino font provided by the mathpazo package.
% Remove the 'osf' bit if you don't like the old style figures.
%\usepackage[T1]{fontenc}
%\usepackage[sc,osf]{mathpazo}

% Set your name here
\def\name{Joseph T. Ornstein}

% Replace this with a link to your CV if you like, or set it empty
% (as in \def\footerlink{}) to remove the link in the footer:
\def\footerlink{https://joeornstein.github.io} 

% The following metadata will show up in the PDF properties
\hypersetup{
  colorlinks = true,
  urlcolor = black,
  pdfauthor = {\name},
  pdfkeywords = {political science, cv},
  pdftitle = {\name: Curriculum Vitae},
  pdfsubject = {Curriculum Vitae},
  pdfpagemode = UseNone
}

\geometry{
  body={6.5in, 8.5in},
  left=1.0in,
  top=1.25in
}

% Customize page headers
\pagestyle{myheadings}
\markright{\name}
\thispagestyle{empty}

% Custom section fonts
\usepackage{sectsty}
\sectionfont{\rmfamily\mdseries\Large}
\subsectionfont{\rmfamily\mdseries\itshape\large}

% Other possible font commands include:
% \ttfamily for teletype,
% \sffamily for sans serif,
% \bfseries for bold,
% \scshape for small caps,
% \normalsize, \large, \Large, \LARGE sizes.

% Don't indent paragraphs.
\setlength\parindent{0em}

% Make lists without bullets
\renewenvironment{itemize}{
  \begin{list}{}{
    \setlength{\leftmargin}{1.5em}
  }
}{
  \end{list}
}

\begin{document}



% Place name at left
%{\huge \name}

% Alternatively, print name centered and bold:
\centerline{\huge \bf \name}

\vspace{0.25in}

\begin{minipage}{0.45\linewidth}
  \href{https://wustl.edu/}{Washington University in St. Louis} \\
  Brown School of Social Work \\
  Hillman Hall \\
  St. Louis, MO 63130\\
\end{minipage}
\begin{minipage}{0.45\linewidth}
  \begin{tabular}{ll}
    %Phone: & (843) 830-2232 \\
    Email: & \href{mailto:jornstein@wustl.edu}{\tt jornstein@wustl.edu} \\
    Homepage: & \href{https://joeornstein.github.io/}{\tt https://joeornstein.github.io/} \\ 
    Google Scholar: & \href{https://scholar.google.com/citations?user=HXC86u8AAAAJ&hl=en}{\tt https://scholar.google.com} \\
    &\\
  \end{tabular}
\end{minipage}

\hrulefill

\section*{Positions}

\begin{itemize}
\item Postdoctoral Research Associate, Washington University in St. Louis, 2018--present.
\item Senior Research Assistant, The Brookings Institution, 2010--2013.
\item Researcher, New England Complex Systems Institute, 2009--2010.
\end{itemize}


\hrulefill

\section*{Education}

\begin{itemize}
  \item Ph.D. Political Science, University of Michigan, 2018.
  	\begin{itemize}
  	\item Committee: Robert Franzese (chair), Jenna Bednar, Elisabeth Gerber, Scott Page
  	\end{itemize}

  \item B.A. Mathematics \& Social Sciences, Dartmouth College, 2009.

  
\end{itemize}

\hrulefill

\section*{Publications}
\begin{enumerate}
\item \href{https://joeornstein.github.io/papers/Ornstein-Hammond2017.pdf}{Ornstein, Joseph T., and Ross A. Hammond. 2017. ``The Burglary Boost: A Note on Detecting Contagion Using the Knox Test.'' \textit{Journal of Quantitative Criminology} 33(1): 65-75.}
\item \href{https://joeornstein.github.io/MonotonicityFailure.html}{Ornstein, Joseph T., and Robert Z. Norman. 2014. ``Frequency of Monotonicity Failure under Instant Runoff Voting: Estimates Based on a Spatial Model of Elections.'' \textit{Public Choice} 161(1-2): 1-9.}
\item \href{https://joeornstein.github.io/papers/Hammond_et_al-2014-Annals_of_the_New_York_Academy_of_Sciences.pdf}{Hammond, Ross A., and Joseph T. Ornstein. 2014. ``A Model of Social Influence on Body Mass Index.'' \textit{Annals of the New York Academy of Sciences} 1331(1): 34-42.}
\item \href{https://joeornstein.github.io/papers/Hammond-2012.pdf}{Hammond, Ross A. et al. 2012. ``A Model of Food Reward Learning with Dynamic Reward Exposure.'' \textit{Frontiers in Computational Neuroscience} 6: 82.}
\end{enumerate}

\hrulefill

\section*{Working Papers}
\begin{enumerate}
\item \href{https://joeornstein.github.io/papers/MLP.pdf}{``Machine Learning and Poststratification'' (revise \& resubmit)} 
\item \href{https://joeornstein.github.io/papers/ornstein-zone-defense.pdf}{``Zone Defense: Why Liberal Cities Build Fewer Houses''}
\item \href{https://joeornstein.github.io/ElectionTiming.html}{``Municipal Election Timing and The Politics of Urban Growth''}
\item \href{https://joeornstein.github.io/papers/dcsm.pdf}{``Diverse Cities Spend More''}
\end{enumerate}

\hrulefill

\section*{Conference Presentations}
\begin{itemize}
\item American Political Science Association (2017)
\item Midwest Political Science Association (2016, 2017)
\item Public Choice Society (2010, 2014)
\end{itemize}

\hrulefill

\section*{Teaching Experience}

\subsection*{Primary Instructor, University of Michigan}
\begin{itemize}
\item POLSCI 369: The Political Economy of Cities  (Summer 2017)
\item Political Science Graduate Student Math Camp (Summer 2016)
\end{itemize}

\subsection*{Graduate Student Instructor, University of Michigan}
\begin{itemize}
\item POLSCI 369: International Economic Relations (with Andrew Kerner) %(Winter 2016)
%	\begin{itemize}
%	\item Professor: Andrew Kerner
%	\end{itemize}
\item POLSCI 391: Introduction to Models (with Scott Page) %(Fall 2015)
%	\begin{itemize}
%	\item Professor: Scott Page
%	\end{itemize}
\item ICPSR: Advanced Empirical Modeling for Theory Evaluation (with Robert Franzese) %(Summer 2015)
%	\begin{itemize}
%	\item Professor: Robert Franzese
%	\end{itemize}
\item ICPSR: Spatial Econometrics Workshop (with Robert Franzese) %(Summer 2015 and 2014)
%	\begin{itemize}
%	\item Professor: Robert Franzese
%	\end{itemize}
\item POLSCI 140: Intro to Comparative Politics (with Robert Franzese) % (Winter 2015)
%	\begin{itemize}
%	\item Professor: Robert Franzese
%	\end{itemize}
\item POLSCI 160: Intro to World Politics (with James Morrow) % (Fall 2014)
%	\begin{itemize}
%	\item Professor: James Morrow
%	\end{itemize}
\item EPID 793: Complex Systems Modeling for Public Health Research (with Ross Hammond) % (July 2015, 2014, and 2013)
%	\begin{itemize}
%	\item Professor: Ross Hammond
%	\end{itemize}
\end{itemize}

\hrulefill

\section*{Awards and Grants}
\begin{itemize}
\item University of Michigan Library Data Grant (2017)
\item Rackham Conference Travel Grant (2013, 2015, 2016, 2017)
\item Dartmouth College Neukom Scholarship (2009)
\end{itemize}

\hrulefill

\section*{Professional Service}

\begin{itemize}
\item \textbf{Reviewer}: \textit{Public Choice}, \textit{Political Science Research and Methods}, \textit{Political Behavior}, \textit{Pediatric Obesity}
%	\begin{itemize}
%	\item \textit{Public Choice}, \textit{Political Behavior}, \textit{Pediatric Obesity}
%	\end{itemize}
\item \textbf{Member}: \textit{American Political Science Association}
%	\begin{itemize}
%	\item American Political Science Association
%	\end{itemize}
\item \textbf{Committees}: University of Michigan Comparative Politics Faculty Search (2017)
\end{itemize}

\hrulefill

\section*{Miscellaneous}
\begin{itemize}
\item \textbf{Software Proficiencies:} Stata, {\tt R}, Python, Java, NetLogo, {\tt LaTeX}
\item \textbf{Erd\H{o}s Number:} 3
\end{itemize}

\bigskip

% Footer
\begin{center}
  \begin{footnotesize}
    Last updated: \today \\
    \href{\footerlink}{\texttt{\footerlink}}
  \end{footnotesize}
\end{center}

\end{document}










