---
title: "Blogdown Features"
author: "Joseph T. Ornstein"
date: '2020-05-10'
slug: blogdown-features
categories: 
 - R
tags: []
images: []
---

Any website built with [Hugo](https://gohugo.io/) and [`blogdown`](https://bookdown.org/yihui/blogdown/) will inevitably feature one blog post about all the cool features of [Hugo](https://gohugo.io/) and [`blogdown`](https://bookdown.org/yihui/blogdown/). This is mine.

Observe, if you will, that we can embed a shiny dashboard into a static website with RMarkdown. Thanks to [Yang Liu](https://liuyanguu.github.io/post/2019/02/24/shiny-in-blogdown/#my-shiny-app-example)!


```{r embed shiny, echo=T}
knitr::include_app("https://vac-lshtm.shinyapps.io/ncov_tracker/?_ga=2.126674445.1246084674.1588877732-1324960155.1586370896", height = "600px")
```

The same `knitr` function also works for slides:

```{r embed slides, echo=T}
knitr::include_url('https://timmastny.rbind.io/slides/first_presentation#2')
```

Here's a tweet courtesy of the Twitter shortcode: 

{{% tweet 877500564405444608 %}}

I will add more examples to this post as I discover them.